Template for indicator documentation

### 2023-09-11 scalingFunctions.Rmd
RenderBook and remove sysDate. Fixed some typos in gjengoing and image widths

Date: `r Sys.Date()`

@@

Define plotting aesthetics
```{r}
low <- "red"
high <- "green"
scale_fill_gradient(low = low, high = high)
ridge <- geom_density_ridges_gradient(scale = 3, size = .3)
myColour <- "black"
mySize <- 10
myAlpha <- .7
myShape <- 21
```

Create random raw data for two variables
```{r}
dat <- data.frame(variable1 = rnorm(50, 10, 4), 
                  variable2 = rnorm(50, 30, 3))

@@

Lets say that the reference condition for variable1 to equal 16, and for variable 2 its 35

@@

To reverse what is good and what is bad condition: change the sign and add one.
```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1*(-1)+1, fill = indicator1*(-1)+1),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+
  geom_vline(xintercept = dat$referenceHigh1[1],
             linetype = "dashed",
             size =2)+

@@

The two indicators become more comparable now that they are normalised. At least this is true around the reference value. But perhaps a variable value around 25 for variabel2 is actually quite bad, and in the above example I have assumed that the value zero is the worst possible value. That may not always be the case.

## Linear, with defined lower limit
We can also say that the worst possible condition for these variables is not zero, but for example say that it is 5 for variable1 and 28 for variable2.
```{r}
dat$referenceLow1 <- 5
dat$referenceLow2 <- 28
```

We can then normalise the data between these two reference values.
```{r}
dat$indicator1_LowHigh <- (dat$variable1-dat$referenceLow1)/(dat$referenceHigh1 - dat$referenceLow1)
dat$indicator1_LowHigh[dat$indicator1_LowHigh <0] <- 0
dat$indicator1_LowHigh[dat$indicator1_LowHigh >1] <- 1

@@

## Break point scaling
We can also define a threshold for good ecological condition. Lets for example say that for variable1, ecosystem function really declines when the variable shows values below 9.

@@

Then we can normalise again, using what I refer to as break point normalisation. For the maths, see [here](https://stats.stackexchange.com/questions/281162/scale-a-number-between-a-range)
```{r}
dat$indicator1_breakPoint <- ifelse(dat$variable1 < dat$thr1,
                                    ((dat$variable1-dat$referenceLow1)/(dat$thr1 - dat$referenceLow1))*(0.6-0)+0,
                                    ((dat$variable1-dat$thr1)/(dat$referenceHigh1 - dat$thr1))*(1-0.6)+0.6
                                    )

@@

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_breakPoint, fill = indicator1_breakPoint),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+
  geom_vline(xintercept = dat$referenceHigh1[1],
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = dat$referenceLow1,
             linetype = "dashed",
             size =2)+
  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+
  geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+

@@

@@

An indicator may have an optimum value somewhere in the middle of its range, rather than a max or min value. You can then define a two-sided indicator.
Lets say that variable2 is optimal when its at a level of 30, and values either lower or high are both bad.

@@

```{r}
dat$indicator2_twoSided <- ifelse(dat$variable2<dat$referenceMid2,
                                  (dat$variable2-dat$referenceLow2)/(dat$referenceMid2-dat$referenceLow2),
                                  ((dat$variable2-dat$referenceMid2)/(dat$referenceHigh2-dat$referenceMid2))*(-1)+1
                                  )

@@

## Sigmoid transformation
There are many other, non-linear ways to normalise an indicator. The correct choice for scaling function varies between indicators and depends on the nature of the indicator and what it represents, but also on the uncertainty around the reference values. Personally, I think the sigmoid function often makes a lot a sense. It is less sensitive to changes around the min and max reference values. A slight deviation from the reference condition can often not be described as a true reduction in condition. This is true for example when reference communities (ecosystemns defines as having or representing good condition) themselves show variation in the variable values, yet the reference value is fixed. A linear scaling will in these cases lead to what is referred to as underestimation biaz (making thing look worse than they really are).

@@

Finding the transformed indicator value when the variable equals the threshold
```{r}
sigmoid_thr <- 100.68*(1-exp(-5*((dat$thr1[1]-dat$referenceLow1[1])/(dat$referenceHigh1[1]-dat$referenceLow1[1]))^2.5))/100

@@

Also plotted against the linearly normalised indicator values:
```{r}
ggplot(dat)+
  geom_point(
    aes(x = indicator1_LowHigh, y = indicator1_sigmoid, fill = indicator1_sigmoid),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+

@@

@@

We can account for the  threshold for good ecological condition when applying sigmoid transformation like this
```{r}
dat$indicator1_sigmoid_thr <- ifelse(dat$variable1<dat$thr1,
                                     ((dat$indicator1_sigmoid-0)/(sigmoid_thr-0))*(0.6-0)+0,
                                     ((dat$indicator1_sigmoid-sigmoid_thr)/(1-sigmoid_thr))*(1-0.6)+0.6
                                     )
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_sigmoid),
             fill   = "grey", 
             colour = myColour,
             size   = 5,
             alpha  = myAlpha,
             shape  = myShape)+

@@

## Exponential functoins
Here are two example of exponential function:

>>>>

March 2023

**Superseeded :** *The functionality explained here is moved over to the [eaTools](https://github.com/NINAnor/eaTools) package. Please see the documentation there for updated examples.*

@@

Define plotting aesthetics

```{r}
low <- "red"
high <- "green"
ridge <- geom_density_ridges_gradient(scale = 3, size = .3)
myColour <- "black"
mySize <- 10
myAlpha <- .7
myShape <- 21
```

Create random raw data for two variables

```{r}
dat <- data.frame(variable1 = rnorm(50, 10, 4), 
                  variable2 = rnorm(50, 30, 3))

@@

Lets say that the reference condition for `variable1` to equal 16, and for variable 2 its 35

@@

To reverse what is good and what is bad condition: change the sign and add one.

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1*(-1)+1, fill = indicator1*(-1)+1),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+
  geom_vline(xintercept = dat$referenceHigh1[1],
             linetype = "dashed",
             size =2)+

@@

The two indicators become more comparable now that they are normalised. At least this is true around the reference value. But perhaps a variable value around 25 for `variable2` is actually quite bad. In the above example I have assumed that the value zero is the worst possible value, but that may not always be the case.

## Linear, with defined lower limit

We can also say that the worst possible condition for these variables is not zero, but for example say that it is 5 for `variable1` and 28 for `variable2`.

```{r}
dat$referenceLow1 <- 5
dat$referenceLow2 <- 28
```

We can then normalise the data between these two reference values.

```{r}
dat$indicator1_LowHigh <- (dat$variable1-dat$referenceLow1)/(dat$referenceHigh1 - dat$referenceLow1)
dat$indicator1_LowHigh[dat$indicator1_LowHigh <0] <- 0
dat$indicator1_LowHigh[dat$indicator1_LowHigh >1] <- 1

@@

## Break point scaling

We can also define a threshold for good ecological condition. Lets for example say that for `variable1`, ecosystem function really declines when the variable shows values below 9.

@@

Then we can normalise again, using what I refer to as break point normalisation. For the maths, see [here](https://stats.stackexchange.com/questions/281162/scale-a-number-between-a-range).

```{r}
dat$indicator1_breakPoint <- ifelse(
  dat$variable1 < dat$thr1,
    ((dat$variable1-dat$referenceLow1)/(dat$thr1 - dat$referenceLow1))*(0.6-0)+0,
    ((dat$variable1-dat$thr1)/(dat$referenceHigh1 - dat$thr1))*(1-0.6)+0.6
  )

@@

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_breakPoint, fill = indicator1_breakPoint),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+
  geom_vline(xintercept = dat$referenceHigh1[1],
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = dat$referenceLow1,
             linetype = "dashed",
             size =2)+
  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
               linetype="solid",
               size=1.5)+
  geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
               linetype="solid",
               size=1.5)+

@@

The solid lines point to the indicator value where the variable equals the threshold value.

@@

An indicator may have an optimum value somewhere in the middle of its range, rather than a max or min value. You can then define a two-sided indicator. Lets say that `variable2` is optimal when its at a level of 30, and values either lower or high are both bad.

@@

```{r}
dat$indicator2_twoSided <- ifelse(
  dat$variable2<dat$referenceMid2,
    (dat$variable2-dat$referenceLow2)/(dat$referenceMid2-dat$referenceLow2),
    ((dat$variable2-dat$referenceMid2)/(dat$referenceHigh2-dat$referenceMid2))*(-1)+1
  )

@@

## Sigmoid transformation

There are many other, non-linear ways to normalise an indicator. The correct choice for scaling function varies between indicators and depends on the nature of the indicator and what it represents, but also on the uncertainty around the reference values. Personally, I think the sigmoid function often makes a lot a sense. It is less sensitive to changes around the min and max reference values. A slight deviation from the reference condition should perhaps not often not be described as a true reduction in condition. This is true for example when reference communities (ecosystems defined as having, or representing, good condition) themselves show variation in the variable values, yet the reference value is fixed. A linear scaling will in these cases lead to what is referred to as underestimation biaz (making thing look worse than they really are).

@@

Finding the transformed indicator value when the variable equals the threshold

```{r}
sigmoid_thr <- 100.68*(1-exp(-5*((dat$thr1[1]-dat$referenceLow1[1])/(dat$referenceHigh1[1]-dat$referenceLow1[1]))^2.5))/100

@@

Also plotted against the linearly normalised indicator values:

```{r}
ggplot(dat)+
  geom_point(
    aes(x = indicator1_LowHigh, y = indicator1_sigmoid, fill = indicator1_sigmoid),
             colour = myColour,
             size   = mySize,
             alpha  = myAlpha,
             shape  = myShape)+

@@

The dotted helper lines selve as visual guides to judge how different a scaled indicator value of 0.6 is for the two approaches.

@@

We can account for the threshold for good ecological condition when applying sigmoid transformation like this

```{r}
dat$indicator1_sigmoid_thr <- ifelse(
  dat$variable1<dat$thr1,
    ((dat$indicator1_sigmoid-0)/(sigmoid_thr-0))*(0.6-0)+0,
    ((dat$indicator1_sigmoid-sigmoid_thr)/(1-sigmoid_thr))*(1-0.6)+0.6
  )
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_sigmoid),
             fill   = "grey", 
             colour = myColour,
             size   = 5,
             alpha  = myAlpha,
             shape  = myShape)+

@@

The grey circles and horizonal line correspons to the sigmoid transformation without any adjustment for the threshold value.

## Exponential functoins

Here are two examples of exponential transformation that can also be used:
